---
output:
  rmarkdown::github_document:
    html_preview: no
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.path = "man/figures/README-", fig.align = "center",
                      fig.height = 4.5, fig.width = 4.5)
```

<!--- README.md is generated from README.Rmd. Please edit that file -->

## aoh: Area of Habitat Maps

[![lifecycle](https://img.shields.io/badge/Lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![R-CMD-check-Ubuntu](https://img.shields.io/github/workflow/status/jeffreyhanson/aoh/Ubuntu/master.svg?label=Ubuntu)](https://github.com/jeffreyhanson/aoh/actions)
[![R-CMD-check-Windows](https://img.shields.io/github/workflow/status/jeffreyhanson/aoh/Windows/master.svg?label=Windows)](https://github.com/jeffreyhanson/aoh/actions)
[![R-CMD-check-Mac-OSX](https://img.shields.io/github/workflow/status/jeffreyhanson/aoh/Mac%20OSX/master.svg?label=Mac%20OSX)](https://github.com/jeffreyhanson/aoh/actions)
[![Coverage Status](https://codecov.io/github/jeffreyhanson/aoh/coverage.svg?branch=master)](https://codecov.io/github/jeffreyhanson/aoh?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/aoh)](https://CRAN.R-project.org/package=aoh)

```{r, include = FALSE}
devtools::load_all()
```

### Overview

TODO.

### Installation

The latest developmental version can be installed using the following _R_ code. This package requires the [_sf_](https://github.com/r-spatial/sf) and [_terra_](https://github.com/r-spatial/terra) _R_ packages which may require additional software to be installed. If you encounter problems installing the _aoh R_ package, please consult the installation instructions for these packages.

```{r, eval = FALSE}
if (!require(remotes)) install.packages("remotes")
remotes::install_github("jeffreyhanson/aoh")
```

### Usage

TODO.

### Citation

Please cite the _aoh R_ package and the underlying datasets used to produce Area of Habitat maps.

```{r, echo  = FALSE, result = "asis", comment = ""}
citation("aoh")
```
