---
output:
  rmarkdown::github_document:
    html_preview: no
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.path = "man/figures/README-", fig.align = "center",
                      fig.height = 4.5, fig.width = 4.5)
```

<!--- README.md is generated from README.Rmd. Please edit that file -->

## aoh: Create Area of Habitat Data

[![lifecycle](https://img.shields.io/badge/Lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![R-CMD-check-Ubuntu](https://img.shields.io/github/workflow/status/prioritizr/aoh/Ubuntu/master.svg?label=Ubuntu)](https://github.com/prioritizr/aoh/actions)
[![R-CMD-check-Windows](https://img.shields.io/github/workflow/status/prioritizr/aoh/Windows/master.svg?label=Windows)](https://github.com/prioritizr/aoh/actions)
[![R-CMD-check-Mac-OSX](https://img.shields.io/github/workflow/status/prioritizr/aoh/Mac%20OSX/master.svg?label=Mac%20OSX)](https://github.com/prioritizr/aoh/actions)
[![Coverage Status](https://codecov.io/github/prioritizr/aoh/coverage.svg?branch=master)](https://codecov.io/github/prioritizr/aoh?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/aoh)](https://CRAN.R-project.org/package=aoh)

```{r, include = FALSE}
devtools::load_all()
```

### Overview

Area of Habitat (AOH) maps aim to delineate the spatial distribution of suitable habitat for a species ([Brooks *et al.* 2019](https://doi.org/10.1016/j.tree.2019.06.009)). They are used to assess performance of protected area systems, measure impacts of threats to biodiversity, and identify priorities for conservation actions (e.g. [Rondinini *et al.* 2005](https://doi.org/10.1111/j.1523-1739.2005.00204.x); [Tracewski *et al.* 2016](https://doi.org/10.1111/cobi.12715); [Durán *et al.* 2021](https://doi.org/10.1111/2041-210X.13427)). These maps are generally produced by obtaining geographic range data for a species, and then removing areas that do not contain suitable habitat or occur outside the known altitudinal limits for the species ([Brooks *et al.* 2019](https://doi.org/10.1016/j.tree.2019.06.009)). To help make these maps accessible, the _aoh R_ package provides routines for automatically creating Area of Habitat data based on the [International Union for Conservation of Nature (IUCN) Red List of Threatened Species](https://www.iucnredlist.org/). After manually downloading species range data from the [IUCN Red List](https://www.iucnredlist.org/resources/spatial-data-download), users can import them (using `read_spp_range_data()`) and then use them to create Area of Habitat data (using `create_spp_aoh_data()`). Global elevation and habitat classification data ([Amatulli *et al.* 2018](https://doi.org/10.1038/sdata.2018.40); [Jung *et al.* 2020](https://doi.org/10.1038/s41597-020-00599-8)) are automatically downloaded, and data on species' habitat preferences and altitudinal limits are obtained automatically using the [IUCN Red List API](https://apiv3.iucnredlist.org/). Since accessing the IUCN Red List requires a token, users may need to [obtain a token](https://apiv3.iucnredlist.org/api/v3/token) and update their _R_ configuration to recognize the token (see installation instructions below for details).

### Installation

#### Package installation

The [latest developmental version of the _aoh R_ package](https://github.com/prioritizr/aoh) can be installed using the following _R_ code. Please note that it requires the [_sf_](https://CRAN.R-project.org/package=sf), [_terra_](https://CRAN.R-project.org/package=terra), and [_zen4R_](https://CRAN.R-project.org/package=zen4R) _R_ packages which may require additional software to be installed. If you encounter problems installing these dependencies, please consult their installation instructions.

```{r, eval = FALSE}
if (!require(remotes)) install.packages("remotes")
remotes::install_github("prioritizr/aoh")
```

#### Accessing the IUCN Red List API

After installing the _aoh R_ package, you will need to obtain a token for the [IUCN Red List API](https://apiv3.iucnredlist.org/) (if you do not have one already). To achieve this, please visit the IUCN website (see <https://apiv3.iucnredlist.org/api/v3/token>) and fill out the form to apply for a token. You should then receive a token shortly after completing the form (but not immediately). After receiving a token, please open the `.Renviron` file on your computer (e.g. using `usethis::edit_r_environ()`). Next, please add the following text to the file (replacing the string with the token) and save the file:

```
IUCN_REDLIST_KEY="your_actual_token_not_this_string"
```

Please restart your R session. You should now be able to access the IUCN Red List API. To verify this, please try running the following _R_ code and -- assuming everything works correctly -- you should see the current version of the IUCN Red List:

```{r, eval = FALSE}
# verify access to IUCN Red List API
rredlist::rl_version()
```

If these instructions did not work, please consult the documentation for the [_rredlist_](https://CRAN.R-project.org/package=rredlist) _R_ package for further details.

### Optional system libraries

The package can leverage the [Geospatial Data Abstraction Library (GDAL)](https://gdal.org/) to help reduce processing time. Although use of GDAL is optional, it can help improve computational performance when not using the preprocessed habitat classification and elevation data. Below we provide platform-specific instructions to install GDAL.

##### _Windows_

The easiest way to install GDAL is through [OSGeo4W](https://www.osgeo.org/). [Download the OSGeo4W installer](https://trac.osgeo.org/osgeo4w/), select the "Express Install" option, and then follow the prompts to complete the installation. After the installation process has finished, please restart your computer. If the _gdalUtils R_ package cannot find GDAL, you may also need to update the `PATH` environmental variable so that it contains the folder path for GDAL (default folder path is `C:\OSGeo4W\bin`),

##### _Ubuntu_

The GDAL binary library need to be installed. For recent versions of Ubuntu (18.04 and later), this library is available through official repositories. It can be installed using the following system commands.

```
apt-get -y update
apt-get install -y gdal-bin
```

##### _Linux_

For Unix-alikes, `gdal` (>= 3.0.2) are required.

##### _MacOS_

The easiest way to install the `gdal` library is using [HomeBrew](https://brew.sh/). After installing HomeBrew, the library can be installed using the following system commands.

```
brew install pkg-config
brew install gdal
```

### Usage

Here we provide a short example for using the _aoh R_ package. In this example, we will generate Area of Habitat data for the following Iberian species: Pyrenean brook salamander (_Calotriton asper_), Iberian frog (_Rana iberica_), western spadefoot toad (_Pelobates cultripes_), and golden striped salamnader (_Chioglossa lusitanica_). Please note that this example is an abridged version of the tutorial provided in the [package vignette](https://prioritizr.github.io/aoh/articles/aoh.html), so please consult the package vignette for a more detailed tutorial on using the package. To start off, we will load the package and several other packages to help with data processing and visualization.

```{r}
# load packages
library(aoh)
library(terra)
library(rappdirs)
library(ggplot2)
```

Now we will import range data for the species. Although users would typically obtain range data from the [International Union for Conservation of Nature (IUCN) Red List of Threatened Species](https://www.iucnredlist.org/), here we will use built-in species range data that distributed with the package for convenience. **Please note that these data were not obtained from the IUCN Red List, and were manually generated using occurrence records from the [Global Biodiversity Information Facility](https://www.gbif.org/).**

```{r}
# find file path for data
path <- system.file("extdata", "EXAMPLE_SPECIES.zip", package = "aoh")

# import data
spp_range_data <- read_spp_range_data(path)

# preview data
print(spp_range_data)
```

Next, we will generate Area of Habitat data for the species. We also specify a folder to cache the downloaded datasets so that we won't need to re-downloaded again during subsequent runs.

```{r, message = FALSE, results = "hide"}
# specify cache directory
cache_dir <- user_data_dir("aoh")

# create cache_dir if needed
if (!file.exists(cache_dir)) {
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
}

# specify folder to save Area of Habitat data
## although we use a temporary directory here to avoid polluting your
## with examples files, you would normally specify the folder
## on your computer where you want to save data
output_dir <- tempdir()

# generate Area of Habitat data
## note that this function might take a complete because it will need to
## download the global habitat and elevation data that first time you run it.
spp_aoh_data <- create_spp_aoh_data(
  spp_range_data, output_dir = output_dir, cache_dir = cache_dir
)

# preview results
print(spp_aoh_data[, c("id_no", "seasonal", "path")])
```

After generating the Area of Habitat data, we can import them.

```{r}
# import the Area of Habitat data
spp_aoh_rasters <- lapply(spp_aoh_data$path, rast)

# preview raster data
print(spp_aoh_rasters)
```

Finally, let's create some maps to compare the range data with the Area of habitat data.

```{r, message = FALSE, warning = FALSE, dpi = 200, fig.width = 5.5, fig.height = 4, out.width = ifelse(isTRUE(knitr:::is_html_output(excludes = c("markdown"))), "60%", "90%")}
# create maps
map <-
  plot_spp_aoh_data(spp_aoh_data, zoom = 6, maptype = "toner-background") +
  scale_fill_viridis_c() +
  scale_color_manual(values = c("range" = "red")) +
  scale_size_manual(values = c("range" = 0.5)) +
  theme(
    axis.title = element_blank(),
    axis.text = element_text(size = 6),
    strip.text = element_text(color = "white"),
    strip.background = element_rect(fill = "black", color = "black")
  )

# display maps
print(map)
```

### Citation

**Please note that the package is still in the early stages of development and not ready for general use. It is not yet suitable for use in scientific publications or related work.**

Please cite the _aoh R_ package and the underlying datasets used to produce Area of Habitat data.

```{r, echo  = FALSE, result = "asis", comment = ""}
citation("aoh")
```

## Getting help

Please refer to the [package website](https://prioritizr.github.io/aoh) for more information. If you have any questions about using the package or suggestions for improving it, please [file an issue at the package's online code repository](https://github.com/prioritizr/aoh/issues).
