---
output:
  rmarkdown::github_document:
    html_preview: no
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "man/figures/README-",
  fig.align = "center",
  fig.height = 4.5,
  fig.width = 4.5,
  dev = "ragg_png"
)
```

<!--- README.md is generated from README.Rmd. Please edit that file -->

## aoh: Create Area of Habitat Data

[![lifecycle](https://img.shields.io/badge/Lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![R-CMD-check-Ubuntu](https://img.shields.io/github/workflow/status/prioritizr/aoh/Ubuntu/master.svg?label=Ubuntu)](https://github.com/prioritizr/aoh/actions)
[![R-CMD-check-Windows](https://img.shields.io/github/workflow/status/prioritizr/aoh/Windows/master.svg?label=Windows)](https://github.com/prioritizr/aoh/actions)
[![R-CMD-check-macOS](https://img.shields.io/github/workflow/status/prioritizr/aoh/macOS/master.svg?label=macOS)](https://github.com/prioritizr/aoh/actions)
[![Coverage Status](https://codecov.io/github/prioritizr/aoh/coverage.svg?branch=master)](https://codecov.io/github/prioritizr/aoh?branch=master)

```{r, include = FALSE}
devtools::load_all()
```

### Overview

Area of Habitat (AOH) maps aim to delineate the spatial distribution of suitable habitat for a species (Brooks *et al.* 2019). They are used to assess performance of protected area systems, measure impacts of threats to biodiversity, and identify priorities for conservation actions (e.g., Rondinini *et al.* 2005; Tracewski *et al.* 2016; Durán *et al.* 2021). These maps are generally produced by obtaining geographic range data for a species, and then removing areas that do not contain suitable habitat or occur outside the known elevational limits for the species (Brooks *et al.* 2019). To help make these maps accessible, the _aoh R_ package provides routines for automatically creating Area of Habitat data based on the [International Union for Conservation of Nature (IUCN) Red List of Threatened Species](https://www.iucnredlist.org/). After manually downloading species range data from the [IUCN Red List](https://www.iucnredlist.org/resources/spatial-data-download), users can import them (using `read_spp_range_data()`), prepare them and collate additional information for subsequent processing (using `create_spp_info_data()`), and then create Area of Habitat data (using `create_spp_aoh_data()`). Global elevation and habitat classification data (Amatulli *et al.* 2018; Jung *et al.* 2020; Lumbierres *et al.* 2021) are automatically downloaded, and data on species' habitat preferences and elevational limits are obtained automatically using the [IUCN Red List API](https://apiv3.iucnredlist.org/). Since accessing the IUCN Red List requires a token, users may need to [obtain a token](https://apiv3.iucnredlist.org/) and update their _R_ configuration to recognize the token (see installation instructions below for details).

### Installation

#### Package installation

The [latest developmental version of the _aoh R_ package](https://github.com/prioritizr/aoh) can be installed using the following _R_ code. Please note that it requires the [_sf_](https://CRAN.R-project.org/package=sf), [_terra_](https://CRAN.R-project.org/package=terra), and [_zen4R_](https://CRAN.R-project.org/package=zen4R) _R_ packages, which may require additional software to be installed. If you encounter problems installing these dependencies, please consult their installation instructions.

```{r, eval = FALSE}
if (!require(remotes)) install.packages("remotes")
remotes::install_github("prioritizr/aoh")
```

#### Optional dependencies

The _aoh R_ package can leverage the _prepr R_ package to augment data cleaning procedures. Since the _prepr R_ package is not available on the Comprehensive R Archive Network (CRAN), it is listed as an optional dependency. In some cases, the _prepr R_ package is required to complete the data cleaning procedures (e.g., to fix especially extreme geometry issues) and the _aoh R_ package will throw an error if the package is not available. To install the _prepr R_ package, please use the following _R_ code. Note that the _prepr R_ package has system dependencies that need to be installed before the package itself can be installed (see below for platform-specific instructions).

```{r, eval = FALSE}
if (!require(remotes)) install.packages("remotes")
remotes::install_github("dickoa/prepr")
```

The package can also leverage the [Geospatial Data Abstraction Library (GDAL)](https://gdal.org/) and [Geographic Resources Analysis Support System (GRASS)](https://grass.osgeo.org/) to help reduce processing time. Although use of these software is optional, they can help improve computational performance when processing data across large spatial extents. Below we provide platform-specific instructions to install dependencies for the _prepr R_ package, GDAL, and GRASS.

##### _Windows_

The [Rtools](https://cran.r-project.org/bin/windows/Rtools/) software needs to be installed to install the _prepr R_ package package. This software provides system requirements from [rwinlib](https://github.com/rwinlib/). Additionally, the easiest way to install GDAL and GRASS is through [OSGeo4W](https://www.osgeo.org/). [Download the OSGeo4W installer](https://trac.osgeo.org/osgeo4w/), select the "Express Install" option, and then follow the prompts to complete the installation. After the installation process has finished, please restart your computer. If the `aoh::is_gdal_python_available()` function cannot find GDAL, you may also need to update the `PATH` environmental variable so that it contains the folder path for GDAL (default folder path is `C:\OSGeo4W\bin`),

##### _Ubuntu_

For recent versions of Ubuntu (18.04 and later), the libraries are available through official repositories. They can be installed using the following system commands.

```
apt-get -y update
apt-get install -y \
  libgdal-dev libgeos-dev libproj-dev gdal-bin grass \
  libgmp3-dev libmpfr-dev libudunits2-dev
```

##### _Linux_

For Unix-alikes, `gdal` (>= 3.0.2), `gmp` (>= 4.2.3), `grass` (>= 7.8.7), and `mpfr` (>= 3.0.0) are required.

##### _MacOS_

The easiest way to install the libraries is using [HomeBrew](https://brew.sh/). After installing HomeBrew, the libraries can be installed using the following system commands.

```
brew tap osgeo/osgeo4mac
brew install pkg-config
brew install gdal
brew install osgeo-grass
brew install gmp
brew install mpfr
```

#### Accessing the IUCN Red List API

After installing the _aoh R_ package, you will need to obtain a token for the [IUCN Red List API](https://apiv3.iucnredlist.org/) (if you do not have one already). To do so, please visit the IUCN API website (<https://apiv3.iucnredlist.org/>), click the "Generate a token" link at the top of the web page, and fill out the form to apply for a token. You should then receive a token shortly after completing the form (but not immediately). After receiving a token, please open the `.Renviron` file on your computer (e.g., using `usethis::edit_r_environ()`). Next, please add the following text to the file (replacing the string with the token) and save the file, using your token in place of the string below.

```
IUCN_REDLIST_KEY="your_actual_token_not_this_string"
```

Please restart your R session. You should now be able to access the IUCN Red List API. To verify this, please try running the following _R_ code and -- assuming everything works correctly -- you should the same output below.

```{r, eval = FALSE}
# verify access to IUCN Red List API
is_iucn_rl_api_available()
```

If these instructions did not work, please consult the documentation for the [_rredlist_](https://CRAN.R-project.org/package=rredlist) _R_ package for further details.


### Usage

Here we provide a short example for using the _aoh R_ package. In this example, we will generate Area of Habitat data for the following Iberian species: Pyrenean brook salamander (_Calotriton asper_), Iberian frog (_Rana iberica_), western spadefoot toad (_Pelobates cultripes_), and golden striped salamnader (_Chioglossa lusitanica_). Please note that this example is an abridged version of the tutorial provided in the [package vignette](https://prioritizr.github.io/aoh/articles/aoh.html), so please consult the package vignette for a more detailed tutorial on using the package. To start off, we will load the package and several other packages to help with data processing and visualization.

```{r, message = FALSE, warning = FALSE}
# load packages
library(aoh)
library(terra)
library(rappdirs)
library(ggplot2)
```

Now we will import range data for the species. Although users would typically obtain range data from the [International Union for Conservation of Nature (IUCN) Red List of Threatened Species](https://www.iucnredlist.org/), here we will use built-in species range data that distributed with the package for convenience. **Please note that these data were not obtained from the IUCN Red List, and were manually generated using occurrence records from the [Global Biodiversity Information Facility](https://www.gbif.org/).**

```{r}
# find file path for data
path <- system.file("extdata", "EXAMPLE_SPECIES.zip", package = "aoh")

# import data
spp_range_data <- read_spp_range_data(path)

# preview data
print(spp_range_data)
```

Next, we will prepare all the range data for generating Area of Habitat data. This procedure -- in addition to repairing any geometry issues in the spatial data --  will obtain information on the species' habitat preferences and elevational limits (via the IUCN Red List of Threatened Species). We also specify a folder to cache the downloaded data so that we won't need to re-download it again during subsequent runs.

```{r, message = FALSE, results = "hide"}
# specify cache directory
cache_dir <- user_data_dir("aoh")

# create cache_dir if needed
if (!file.exists(cache_dir)) {
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
}

# prepare information
spp_info_data <- create_spp_info_data(spp_range_data, cache_dir = cache_dir)
```

We can now generate Area of Habitat data for the species. By default, these data will be generated using elevation data derived from Amatulli *et al.* (2018) and habitat data derived from Lumbierres *et al.* (2021). Similar to before, we also specify a folder to cache the downloaded datasets so that we won't need to re-downloaded again during subsequent runs.

```{r, message = FALSE, results = "hide"}
# specify cache directory
cache_dir <- user_data_dir("aoh")

# specify folder to save Area of Habitat data
## although we use a temporary directory here to avoid polluting your
## with examples files, you would normally specify the folder
## on your computer where you want to save data
output_dir <- tempdir()

# generate Area of Habitat data
## note that this function might take a complete because it will need to
## download the global habitat and elevation data that first time you run it.
spp_aoh_data <- create_spp_aoh_data(
  spp_info_data, output_dir = output_dir, cache_dir = cache_dir
)

# preview results
print(spp_aoh_data[, c("id_no", "seasonal", "path")])
```

After generating the Area of Habitat data, we can import them.

```{r}
# import the Area of Habitat data
spp_aoh_rasters <- lapply(spp_aoh_data$path, rast)

# preview raster data
print(spp_aoh_rasters)
```

Finally, let's create some maps to compare the range data with the Area of habitat data.

```{r "map", message = FALSE, warning = FALSE, dpi = 200, fig.width = 5.5, fig.height = 4, out.width = ifelse(isTRUE(knitr::is_html_output(excludes = c("markdown"))), "60%", "90%")}
# create maps
## N.B. you might need to install the ggmap package
map <-
  plot_spp_aoh_data(spp_aoh_data, zoom = 6, maptype = "toner-background") +
  scale_fill_viridis_d() +
  scale_color_manual(values = c("range" = "red")) +
  scale_size_manual(values = c("range" = 0.5)) +
  theme(
    axis.title = element_blank(),
    axis.text = element_text(size = 6),
    strip.text = element_text(color = "white"),
    strip.background = element_rect(fill = "black", color = "black")
  )

# display maps
print(map)
```

### Citation

Please cite the _aoh R_ package and the underlying datasets used to produce Area of Habitat data.

```{r, echo  = FALSE, result = "asis", comment = ""}
citation("aoh")
```

## Getting help

Please refer to the [package website](https://prioritizr.github.io/aoh) for more information. If you have any questions about using the package or suggestions for improving it, please [file an issue at the package's online code repository](https://github.com/prioritizr/aoh/issues).
