---
title: "Customization"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: true
    self_contained: yes
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
vignette: >
  %\VignetteIndexEntry{Customization}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
---

## Introduction

The _aoh R_ package can be used to generate Area of Habitat data using various datasets. By default, it will create Area of Habitat data using elevation data derived from Robinson _et al._ [-@r4] and habitat classification data derived from Lumbierres _et al._ [-@r9]. In addition to these default datasets, it also provides built-in functions to use habitat classification derived from other data sources [e.g., from @r5]. It can also be used to generate Area of Habitat data using elevation and habitat classification data that has been manually generated (e.g., from [Copernicus Corine Land Cover](https://land.copernicus.eu/pan-european/corine-land-cover)). As such, the package provides flexible framework for generating customized Area of Habitat data.

## Tutorial

Here we will show how Area of Habitat data can be generated using manually specified elevation, habitat classification, and crosswalk data. If you have not previously read through the [Getting started vignette](aoh.html), this is strongly recommended to provide an introduction to the package. To start off, we will load the package. We will also load the [_rappdirs_](https://CRAN.R-project.org/package=rappdirs) _R_ package to cache data, and the [_terra_](https://CRAN.R-project.org/package=terra) and [_ggplot2_](https://CRAN.R-project.org/package=ggplot2) _R_ packages to visualize results.

```{r, message = FALSE, warning = FALSE}
# load packages
library(aoh)
library(terra)
library(rappdirs)
library(ggplot2)
```

Now we will import range data for the species. Although users would typically obtain range data from the [International Union for Conservation of Nature (IUCN) Red List of Threatened Species](https://www.iucnredlist.org/), here we will use built-in species range data that distributed with the package for convenience. **Please note that these data were not obtained from the IUCN Red List, and were manually generated using occurrence records from the [Global Biodiversity Information Facility](https://www.gbif.org/).**

```{r}
# find file path for data
path <- system.file("extdata", "EXAMPLE_SPECIES.zip", package = "aoh")

# import data
spp_range_data <- read_spp_range_data(path)

# preview data
print(spp_range_data)
```

Next, we will import elevation data. Specifically, we will import a pre-processed version of the Robinson _et al._ [-@r4] dataset. If you do have not  this datset data already on your computer, it will be automatically downloaded.

```{r, fig.height = 4, fig.width = 8}
# specify cache directory
cache_dir <- user_data_dir("aoh")

# create cache_dir if needed
if (!file.exists(cache_dir)) {
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
}

# import elevation data
elevation_data <- get_global_elevation_data(dir = cache_dir)

# preview data
print(elevation_data)

# visualize data
plot(elevation_data, main = "Elevation data")
```

```{r, eval = FALSE}
# print file path for where elevation data is stored on disk
print(sources(elevation_data))
```

```{r, echo = FALSE}
print(gsub(Sys.getenv("USER"), "user", sources(elevation_data), fixed = TRUE))
```

```{r}
# although we used a function to automatically find and import the elevation
# data, we could use the following code to import the data using a file path:
#> elevation_data <- rast("insert_file_path_here")
```

Next, we will import habitat classification data. Similar to the elevation data, we will import a pre-processed version of the habitat classification data. Specifically, we will import a pre-processed version of the level 1 habitat classification produced by Jung *et al.* [-@r5].

```{r, fig.height = 4, fig.width = 8}
# import elevation data
habitat_data <- get_jung_lvl1_habitat_data(dir = cache_dir)

# preview data
print(habitat_data)

# visualize data
plot(habitat_data, main = "Habitat classification data")
```

```{r, eval = FALSE}
# print file path for where habitat data is stored on disk
print(sources(habitat_data))
```

```{r, echo = FALSE}
print(gsub(Sys.getenv("USER"), "user", sources(elevation_data), fixed = TRUE))
```

```{r}
# although we used a function to automatically find and import the habitat
# data, we could use the following code to import the data using a file path:
#> habitat_data <- rast("insert_file_path_here")
```

Critically, the elevation data and habitat classification data must have exactly the same spatial properties. This means they must have the coordinate reference system, resolution, and spatial extent. If you are using elevation or habitat classification data that you have previously prepared yourself (or manually download from online sources), you may need to resample (or reproject) your data to ensure both datasets have the same spatial properties using a geographic information system (GIS). For example, data could be resampled using [ESRI ArcGIS](https://www.esri.com/en-us/arcgis/about-arcgis/overview), [QGIS](https://qgis.org/en/site/), [terra R package](https://cran.r-project.org/package=terra), or the [gdalUtilities R package](https://cran.r-project.org/package=gdalUtilities)). We can verify that both of the elevation and habitat classification datasets have the same spatial properties using the following code.

```{r}
# verify that elevation and habitat classification data have the same
# spatial properties
compareGeom(elevation_data, habitat_data)

# if this code returns TRUE, this means they have the same spatial properties.
# otherwise, if we see an error, then this means that they do not have
# have the same spatial properties and require updating
```

We will now import a crosswalk table for the habitat classification data. A crosswalk table specifies which pixel values in the habitat classification data correspond to which [IUCN habitat classes](https://www.iucnredlist.org/resources/habitat-classification-scheme). This table can specify one-to-one relationships (e.g., pixel value 12 corresponds to IUCN class 1.1), one-to-many relationships (e.g., pixel values 12 corresponds to IUCN classes 1.1 and 1.2), and many-to-many relationships (e.g., both pixel values 12 and 13 each correspond to IUCN classes 1.1 and 1.2). Since the package provides a built-in crosswalk table for level 1 habitat classification produced by Jung *et al.* [-@r5], we will use these data. If you wish to use a different crosswalk table, please ensure that it follows the same format (i.e., same column names and data types).

```{r}
# import crosswalk table
data(crosswalk_jung_lvl1_data)

# assign table to a new
crosswalk_data <- crosswalk_jung_lvl1_data

# print table
## code column contains codes for the IUCN habitat classes
## value column contains values in the habitat classification data
print(crosswalk_data, n = Inf)

# although we used a function to automatically import the crosswalk table,
# we could use the following code to import the data from a CSV file
#> crosswalk_data <- readr::read_csv("insert_file_path_here")
```

After importing all the data, we can now generate the Area of Habitat data.

```{r, message = FALSE, results = "hide"}
# specify folder to save Area of Habitat data
## although we use a temporary directory here to avoid polluting your
## with examples files, you would normally specify the folder
## on your computer where you want to save data
output_dir <- tempdir()

# prepare information
spp_info_data <- create_spp_info_data(spp_range_data, cache_dir = cache_dir)

# generate Area of Habitat data
## note that we specify exactly which elevation, habitat classification,
## and crosswalk data should be used for processing
spp_aoh_data <- create_spp_aoh_data(
  spp_info_data,
  elevation_data = elevation_data,
  habitat_data = habitat_data,
  crosswalk_data = crosswalk_data,
  output_dir = output_dir,
  cache_dir = cache_dir
)
```

```{r}
# preview results
## resulting dataset is a simple features (sf) object containing
## spatial geometries for cleaned versions of the range data
## (in the geometry column) and the following additional columns:
##
## - id_no            : IUCN Red List taxon identifier
## - seasonal         : integer identifier for seasonal distributions
## - category         : character IUCN Red List threat category
## - full_habitat_code: All IUCN Red List codes for suitable habitat classes
##                      (multiple codes are delimited using "|" symbols)
## - habitat_code     : IUCN Red List codes for suitable habitat classes
##                      used to create AOH maps
## - elevation_lower  : lower limit for the species on IUCN Red List
## - elevation_upper  : upper limit for the species on IUCN Red List
## - xmin             : minimum x-coordinate for Area of Habitat data
## - xmax             : maximum x-coordinate for Area of Habitat data
## - ymin             : minimum y-coordinate for Area of Habitat data
## - ymax             : maximum y-coordinate for Area of Habitat data
## - path             : file path for Area of Habitat data (GeoTIFF format)
##
## since data obtained from the IUCN Red List cannot be redistributed,
## we will not show all the columns in this object
##
## N.B. you can view all columns on your computer with:
##> print(spp_aoh_data, width = Inf)
print(spp_aoh_data[, c("id_no", "binomial", "seasonal", "path")])
```

Finally, let's create some maps to compare the range data with the Area of habitat data. Although we could create these maps manually (e.g., using the [_ggplot2_](https://CRAN.R-project.org/package=ggplot2) _R_ package), we will use a plotting function distributed with the _aoh R_ package for convenience.

```{r, message = FALSE, warning = FALSE, dpi = 200, fig.width = 5.5, fig.height = 4, out.width = "90%"}
# create maps
map <-
  plot_spp_aoh_data(spp_aoh_data, zoom = 6, maptype = "toner-background") +
  scale_fill_viridis_d() +
  scale_color_manual(values = c("range" = "red")) +
  scale_size_manual(values = c("range" = 0.5)) +
  theme(
    axis.title = element_blank(),
    axis.text = element_text(size = 6),
    strip.text = element_text(color = "white"),
    strip.background = element_rect(fill = "black", color = "black")
  )

# display maps
print(map)
```

## Conclusion

Hopefully, this vignette has provided a useful overview for customizing Area of Habitat data. If you have any questions or suggestions for additional elevation, habitat classification, or crosswalk datasets that could be included in the package, please [file an issue at the package's online code repository](https://github.com/prioritizr/aoh/issues).

## References
